---
title: "Introducing geniusr"
date: "2017-11-09"
type: "post"
tags: ["R", "Package", "Genius", "API"]
---

```{r setup, include=FALSE}
# markdown setup
knitr::opts_chunk$set(cache=TRUE, echo = FALSE, tidy = TRUE, warning = FALSE, message = FALSE, out.width = '100%', dpi = 180)

# load pkgs
# devtools::install_github("ewenme/geniusr")
library(geniusr)
library(tidyverse)
library(tidytext)
```

I made a new R package (`r emo::ji("fire")` `r emo::ji("package")` `r emo::ji("fire")`) this week. My [sophomore post here](https://ewenme.rbind.io/blog/2017-03-20-kanye_sentiment_analysis/) (from way back in March) was a text-based analysis of Kanye West records, which exposed me to the [Genius](https://genius.com/) API[^documentation] treasure trove once I wanted to get hold of lyrical content. In the process, I'd written some functions for playing nicely with this API in R that would probably be handy for others dabbling/wanting to dabble in similar matters. Since then, I've found a window of time to build on and consolidate this work into a well-rounded package. `geniusr` is now available on CRAN (with the dev version sat on GitHub [here](https://github.com/ewenme/geniusr). Time for the tour.[^admin] 

## Searchin'

The Genius API's search capabilities allows for users to scan all hosted content on the site (i.e. all songs). A couple of quick implementations exist in the package for retrieving artist (`search_artist()`) and song (`search_song()`) matches for search terms. Let's start by looking for songs brought back by a simple search for "Hennessey".

```{r song_search}
search_song(search_term = "Hennessey", n_results = 50)
```

## Goin' Meta

Wanna bring back some fuller information about a particular song (`get_song_meta()`), artist (`get_artist_meta()`) or album (`get_album_meta()`)? Well, now you can. Let's test one out - how about a Hennessey-themed track from our last search.

```{r song_meta}
get_song_meta(song_id = 6596) %>% glimpse
```


