---
title: "ewenthemes (AKA modding hrbrthemes)"
description: "Achieve ggplot + website zen."
date: '2018-11-05'
slug: ewenthemes-aka-modding-hrbrthemes
tags: ["r", "ggplot2", "typography"]
draft: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE, warning = FALSE, message = FALSE, echo = FALSE, out.width = '75%', dev = 'svg', fig.align='center')
```

This post is a short one, (partly) accompanying the last one on [typography for blogodown](https://ewen.io/2018/11/01/readable-accessible-typography-blogdown/). 

If you use [ggplot2](https://ggplot2.tidyverse.org/), you might want to make sure your site's typeface(s) are mirrored in your featured plotting handywork. You could also take this opportunity to develop some custom `ggplot2` theme fluorishes, helping your plots and posts make a charming fit.

[Bob Rudis](https://rud.is/b/) (AKA [hrbrmstr](https://twitter.com/hrbrmstr)) made perhaps *the* seminal ggplot2 themes package so far, [hrbrthemes](https://hrbrmstr.github.io/hrbrthemes/). [Bob nicely explains](https://hrbrmstr.github.io/hrbrthemes/articles/why_hrbrthemes.html) the opinions that form "what's in the box", chiefly a desire for good typography and minimal chart junk. It's really nice.

```{r, out.width="100%"}
library(ggplot2)
library(hrbrthemes)

ggplot(mtcars, aes(mpg, wt)) +
  geom_point() +
  labs(x="Fuel efficiency (mpg)", y="Weight (tons)",
       title="Seminal ggplot2 scatterplot example",
       subtitle="A plot that is only useful for demonstration purposes",
       caption="Brought to you by the letter 'g'") + 
  # apply theme
  theme_ipsum()
```

So, I made my own twist of a classic - introducing [ewenthemes](https://github.com/ewenme/ewenthemes). I plan to use/maintain the package to create charts that sit beautifully within posts on the site.

```{r, out.width="100%"}
library(ggplot2)
library(ewenthemes)

ggplot(mtcars, aes(mpg, wt)) +
  geom_point() +
  labs(x="Fuel efficiency (mpg)", y="Weight (tons)",
       title="Seminal ggplot2 scatterplot example",
       subtitle="A plot that is only useful for demonstration purposes",
       caption="Brought to you by the letter 'g'") + 
  # apply theme
  theme_ewen_ws()
```

Following Bob's lead, I ship [Work Sans](https://github.com/weiweihuanghuang/Work-Sans) inside the software and include functions to install it on your system with help from `extrafont`. Font handling is surprisingly tricksy in R and this seems to be the best option right now.

The current core themes, `theme_ewen` (Arial Narrow-based) and `theme_ewen_ws` (Work Sans-based) draw heavily from `hrbrthemes`' `theme_ipsum` with some modifications - the chart grid runs on major horizontal lines only, and a thick x-axis line is used by default. Some text properties, notably size and colour, are also modified for better 
