---
title: "Introducing geniusr"
description: "I made an R interface to the Genius API."
date: "2017-12-06"
tags: ["r", "package", "genius", "geniusr"]
draft: no
---

```{r setup, include=FALSE}
# markdown setup
knitr::opts_chunk$set(cache=TRUE, echo = TRUE, warning = FALSE, message = FALSE, dev = 'svg')

# load pkgs
library(geniusr)
library(dplyr)
library(tidytext)
library(ggplot2)
library(waffle)
library(extrafont)
```

I made a new R package (`r emo::ji("fire")``r emo::ji("package")``r emo::ji("fire")`) this week. My [sophomore post on this site](https://ewenme.rbind.io/blog/2017-03-20-kanye_sentiment_analysis/) (from way back in March) was a text-based analysis of Kanye West records, which exposed me to the [Genius](https://genius.com/) API[^documentation] once I wanted to get hold of some lyrical content. In the process, I wrote some functions for playing nicely with this API in R that I knew would probably be handy for others dabbling/wanting to dabble in similar matters.

Since then, I've found a window of time to build on and consolidate this work into a well-rounded package. `geniusr` is now [available on CRAN](https://cran.r-project.org/web/packages/geniusr/index.html). This is the tour.[^admin] 

## Searchin'

The Genius API's search capabilities allows for users to scan all hosted content on the site (i.e. all songs). A couple of quick implementations exist in the package for retrieving artist (`search_artist()`) and song (`search_song()`) matches for search terms. Let's start by looking for songs brought back by a simple search for "Versace".

```{r song_search}
versace_search <- search_song(search_term = "Versace") %>% glimpse
```

## Goin' Meta

Wanna bring back some fuller information about a particular song (`get_song_meta()`), artist (`get_artist_meta()`) or album (`get_album_meta()`)? Well, now you can. Let's test one out - how about investigating a classic Versace-themed track from our last search, 'Versace' by Migos.

```{r song_meta}
versace_meta <- versace_search %>% 
  filter(artist_name == "Migos", song_name == "Versace") %>% 
  pull(song_id) %>% 
  get_song_meta() %>% 
  glimpse
```

With this metadata, we can easily pull up the song's artwork, for example. Why not.

```{r artwork, out.width = "75%", fig.align = "center"}
artwork <- versace_meta$song_art_image_url

knitr::include_graphics(path = artwork)
```

## Gettin' Lyrical

The package allows you to grab lyrics for tracks real quick, using the `scrape_lyrics_id()` (or `scrape_lyrics_url()`) function.[^scrape] While we're at it, let's call on `tidytext` and answer a question for the people - *just how much were the Migos chatting about "Versace" in 'Versace'?*

```{r lyrics, warning=FALSE}
# get words proportion representation
versace_prop <- scrape_lyrics_url(versace_meta$song_lyrics_url) %>%
  unnest_tokens(word, line) %>%
  count(word) %>%
  mutate(proportion = n / sum(n)) %>%
  # was the word versace or the rest
  mutate(versace_or_the_rest=if_else(word == "versace", "Versace", "Not Versace")) %>%
  group_by(versace_or_the_rest) %>%
  summarise(proportion=sum(proportion)*100)

# waffle plot
parts <- c("Not Versace" = versace_prop$proportion[[1]],
           "Versace" = versace_prop$proportion[[2]])

waffle(parts, rows = 8, size = 1, colors = c("#3C3A32", "#E3BD08"), use_glyph = "volume-up", 
       glyph_size = 10, title = "How much of the Migos - 'Versace' lyrics are, indeed, Versace?") +
  labs(caption="@ewen_") +
  theme_void(base_family = "Work Sans", base_size = 12) +
  theme(legend.position = "bottom")
```

The word "Versace" represents 32.5% of the lyrics in Migos - Versace. Phew.

That's all for now, enjoy the package and keep up with future developments over at the [GitHub repo](https://github.com/ewenme/geniusr).

[^documentation]: Genius' [documentation pages](https://docs.genius.com/) are helpful if you want to know more about the API.
[^admin]: For installation & authentication guidance, head for the [README](https://github.com/ewenme/geniusr). 
[^scrape]: As well as API helper functions, the package includes things that are scraping the Genius site (Genius' API doesn't include lyrics, for example). Be sure to read the [terms of service](https://genius.com/static/terms) in case of any concerns.
