---
title: 'Boy In Da Centre: A network analysis of UK Grime'
author: ''
date: '2021-09-20'
slug: boy-in-da-centre-a-network-analysis-of-uk-grime
categories: []
tags:
  - Lorem
  - Ipsum
description: Sed ut perspiciatis unde omnis iste natus error sit voluptatem
draft: true
---

```{r setup, include=FALSE}
library(jsonlite)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tidygraph)
library(ggraph)
library(shadowtext)
library(visNetwork)
library(ggplot2)

# get datasets
releases <- jsonlite::read_json(
  "https://raw.githubusercontent.com/club-cooking/grime-archives/master/data/releases.json", 
  simplifyVector = TRUE
  )

# relations_df <- igraph::as_data_frame(relations)
# 
# filter_graph <- function(x, min_year, max_year, min_have = 20, styles = NULL) {
# 
#   df <- igraph::as_data_frame(x)
# 
#   if (is.null(styles)) {
#     df_subset <- df %>%
#       dplyr::filter(
#       between(year, min_year, max_year),
#       community_have >= min_have
#       )
#   } else {
#     df_subset <- df %>%
#       dplyr::filter(
#       between(year, min_year, max_year),
#       community_have >= min_have,
#       style %in% styles
#       )
#   }
# 
#   as_tbl_graph (df_subset) %>%
#     activate(nodes) %>%
#     mutate(degree = centrality_degree()) %>%
#     activate(edges) %>%
#     mutate(centrality = centrality_edge_betweenness()) %>%
#     arrange(centrality)
# }
```

Grime isn't a pre internet genre of music, but it is pre social media. It is rooted in specific postcodes and working-class, Black British beginnings:

> Grime music, the genre which took the British charts by storm is deeply connected to Roman Road and the surrounding area. The sound which arguably defines urban Britain in the noughties was honed and crafted by the likes of Wiley and Dizzee Rascal on the street corners and in the tower blocks of Bow and the Roman. Rhythm Division, the now infamous record shop run by Risky Roadz, was a hub of activity for the artists and was the scene of countless freestyle battles and mic-offs.

Source: [Key grime figures and their roots with Roman Road — Roman Road LDN](https://romanroadlondon.com/key-grime-figures-history-roman-road/) by @romanroadldn

The scene's origins were characterised by grass-roots interdependence, a "Do It Together" ethos. Localised groups of like-minded MCs and DJs collectivised as "crews". Radio stations would be at least part-funded by "subs" paid by the roster. Producers would press up runs of white label 12"s themselves, and shift them to record store buyers the same day:

> Many producers bring white labels of their latest tunes fresh from the pressing plant and sell them exclusively at Rhythm Division before distributing them elsewhere. A hot tune will have been a staple on the pirates and at raves for months before its official release and may shift a few hundred copies in just a week. Wiley claims to have sold up to 50,000 last year out of the back of his car.

Source: [A look at grime music](https://www.theguardian.com/music/2004/may/23/urban1) by @guardian

Technology of the time was subverted to (literally, at times) signal boost the scene. FM transmitters were [ran up tower blocks](https://www.standard.co.uk/hp/front/asbo-bars-pirate-dj-from-the-rooftops-7173468.html) to broadcast pirate radio sets beyond the borough they started in. Bluetooth-enabled mobile phones helped 96kbps vinyl rips ripple through school playgrounds. Making these connections took effort and sacrifice, but led to intimate bonds between Grime's participants and punters.

The activities described above paint a picture of a vibrant IRL social network, that is inseparable from the artistic practice itself. From claustrophobic radio sets to classic crew records, Grime is stacked with collaborative moments that made the product more than the sum of it's parts. This post investigates how the highly-networked Grime scene can be mapped out visually with data from [Discogs](https://www.discogs.com/) alone. Discogs doesn't provide a complete archive of recorded Grime (Discogs is crowd-sourced and biased towards music made available on vinyl formats), but can act as a solid proxy for understanding ever-changing relationships between Grime artists and give a sense of the scene's "health".

## Understanding the data

The dataset consists of 4,290 releases between 2000 and 2020 with Discogs entries.

```{r}
releases %>% 
  dplyr::filter(!is.na(year)) %>% 
  ggplot(aes(x = year)) +
  geom_histogram(stat = "count")
```


## Origins, 2000-02

Up until 2002, in terms of released music, the Grime scene was just a few small pockets of activity. One central figure was already emerging...Wiley.

```{r origins, eval=FALSE}
origins <- filter_graph(relations, min_year = 2000, max_year = 2002, min_have = 0)

ggraph(origins, layout = 'fr') + 
  geom_edge_fan(
    aes(
      start_cap = label_rect(node1.name), end_cap = label_rect(node2.name))
    ) + 
  geom_shadowtext(
    aes(label=name, x=x, y=y), 
    family = "Recursive Regular", bg.r = 0.1, size = 4.25
    ) +
  theme_graph(title_family = "Recursive Bold", subtitle_family = "Recursive Light",
              caption_family = "Recursive Light") +
  labs(title = "Recorded Grime: origins, 2000-02",
       subtitle = "Connections = collaborations on records listed on Discogs under Style: Grime",
       caption = "@ewen_")
```

If you hadn't already guessed, Wiley was well-established with direct connections to a bunch of artists. He had an unmistakeable creative purple patch with Flow Dan, collaborating on four separate records in this time. Honourable mentions should also go to Geeneus and Terror Danjah, tied for the 2^nd most connected individual across this formative period.

```{r echo=FALSE}
blogdown::shortcode('youtube', 'rwJEVr9F6hg')
```

## Early development, 2003-04

2003 was definitely a break-out year for the genre, as Dizzee Rascal won a Mercury Prize and Channel U landed on satellite TV. 2004 saw Lethal Bizzle crash the top ten UK single charts twice through "Oi!" with More Fire Crew and "Pow! (Forward)" alone.

```{r development, eval=FALSE}
development <- filter_graph(
  relations, min_year = 2003, max_year = 2004, styles = "Grime"
  )

ggraph(development, layout = 'fr') + 
  geom_edge_fan(
    aes(
      start_cap = label_rect(node1.name), end_cap = label_rect(node2.name))
  ) +
  geom_shadowtext(
    aes(label=name, x=x, y=y, size=degree),
    family = "Recursive Regular", bg.r = 0.2, show.legend = FALSE, check_overlap = TRUE
  ) +
  scale_size_continuous(range = c(2.5, 5)) +
  theme_graph(title_family = "Recursive Bold", subtitle_family = "Recursive Light",
              caption_family = "Recursive Light") +
  labs(title = "Recorded Grime: Development, 2003-04",
       subtitle = "Connections = collaborations on records listed on Discogs under Style: Grime",
       caption = "@ewen_")
```

The scene's growth is immediately visible from this network graph. The two artists mentioned above are central figures, and most of the scene is directly or indirectly connected. Wiley remains a key link-up between camps.

There are still some interesting outliers that suggest a healthy injection of new energy was occuring. Ruff Sqwad began a prolific run of white labels in this period, and it's members appear to have been almost entirely insulated from the rest of the scene (in recorded music, at least) during their formative years. To this day, the Ruff Sqwad catalogue is out on it's own.

```{r echo=FALSE}
blogdown::shortcode('youtube', 'pMDHjMagG0Y')
```

## Crossover, 2005-06

If 2003 was the year Grime was getting looks from the mainstream, 2005 was the year of the cross-over. Roll Deep, a collective featuring early scene-builders Wiley and Riko Dan, released a full-length "In At The Deep End" that dabbled in more "commercial" sounds.

```{r echo=FALSE}
blogdown::shortcode('youtube', 'wjREW-QULyQ')
```

Despite this new venture, the group was still collaborating within the wider scene. Conversely, a couple of high-profile artists that emerged during this period came from outside of the ecosystem entirely.

```{r crossover, eval=FALSE}
crossover <- filter_graph(
  relations, min_year = 2005, max_year = 2006, styles = "Grime", min_have = 50
)

ggraph(crossover, layout = 'fr') + 
  geom_edge_fan(
    aes(
      start_cap = label_rect(node1.name), end_cap = label_rect(node2.name))
  ) +
  geom_shadowtext(
    aes(label=name, x=x, y=y, size=degree),
    family = "Recursive Regular", bg.r = 0.2, show.legend = FALSE, check_overlap = TRUE
  ) +
  scale_size_continuous(range = c(2, 5))  +
  theme_graph(title_family = "Recursive Bold", subtitle_family = "Recursive Light",
              caption_family = "Recursive Light") +
  labs(title = "Recorded Grime: Cross-over, 2005-06",
       subtitle = "Connections = collaborations on records listed on Discogs under Style: Grime",
       caption = "@ewen_")
```

In 2006, Lady Sovereign signed to Roc-A-Fella Records and gained access to household names like Missy Elliot in the process. At roughly the same time, Mike Skinner 

## Splinters, 2007-08

...

```{r splinters, eval=FALSE}
splinters <- filter_graph(
  relations, min_year = 2007, max_year = 2008, styles = "Grime", min_have = 50
)

ggraph(splinters, layout = 'fr') + 
  geom_edge_fan(
    aes(
      start_cap = label_rect(node1.name), end_cap = label_rect(node2.name))
  ) +
  geom_shadowtext(
    aes(label=name, x=x, y=y, size=degree),
    family = "Recursive Regular", bg.r = 0.2, show.legend = FALSE, check_overlap = TRUE
  ) +
  scale_size_continuous(range = c(2, 5))

```

...

## Faithful, 2009-10

...

```{r faithful, eval=FALSE}
faithful <- filter_graph(
  relations, min_year = 2007, max_year = 2008, styles = "Grime", min_have = 50
)

ggraph(faithful, layout = 'fr') + 
  geom_edge_fan(
    aes(
      start_cap = label_rect(node1.name), end_cap = label_rect(node2.name))
  ) +
  geom_shadowtext(
    aes(label=name, x=x, y=y, size=degree),
    family = "Recursive Regular", bg.r = 0.2, show.legend = FALSE, check_overlap = TRUE
  ) +
  scale_size_continuous(range = c(2, 5))

```
